---
title: "The cheby package (with applications in economics)"
author: "Philip Barrett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The cheby package (with applications in economics)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction

The `cheby` package provides easy-to-use Chebychev interpolation of an arbitrary R function.

Chebychev interpolation approximates an arbitrary function on a closed interval using an the Chebychev polynomials as a basis.  In general, polynomial approximation is helpful when a function $f:[a,b]\rightarrow \mathbb R$ is expensive to compute at all values in the domain.  So instead, one approximates $f(\cdot)$ is by $\hat f(\cdot)$, where:

  $$\hat f(x) = \sum_{i=1}^n a_i \phi_i(x) $$

Where $\phi_i(x)$ are polynomials.  In practice, this involves computing the function at a limited number of grid points, and then fitting a polynomial function through the function values at these points.  This fitting is typically done by linear regression of the points on the fitting polynomials.  This minimizes the square deviations at the grid points, but imposes no restrictions on the shape of the approximation.

If we want the resulting approximation to have particular properties (such as monotonicity or concavity), then we will need to use other methods.  This package provides functionality to fit one-dimensional functions both via convential Chebychev approximation, and by imposing constraints on the shape of the approximation.

The algorithms in this package draw heavily on Judd (1998).

# One-dimensional Chebychev approximation

## The Theory

Algorithm 6.2 of Judd (1998) outlines the steps required to produce the one dimensional Chebychev interpolation of order $n$ on $m$ nodes (with $n<m$) of the function $f:[a,b]\rightarrow \mathbb R$.

1. Compute the interpolaton nodes on $[-1,1]$ via:
    $$z_k = -\cos \left( \frac{2k-1}{2m}\pi \right)$$
1. Map the nodes to the interval $[a,b]$
    $$x_k = \left( \frac{b-a}{2} \right)(z_k+1) + a$$
1. Compute $f(\cdot)$ at the nodes
    $$y_k = f(x_k), \qquad k=1,\ldots,m$$
1. Regress the $y_k$ on the Chebychev polynomials $T_i(z)$ for $i=0,\ldots,n$
    $$a_i = \frac{\sum_{k=1}^m y_k T_i(z_k) }{\sum_{k=1}^m T_i(z_k)^2}$$
1. The Chebychev approximation is then given by:
    $$\hat f(x) = \sum_{i=1}^n a_i T_i\left(\frac{2(x-1)}{b-a}-1\right)$$

Theorem 6.7.3 of Judd (1998) shows that this approximation is pointwise-convergent for any $C^k$ function.

## Basic approximation

The basic command for Chebychev interpolation is `d1.poly`.  For example, to compute the approximation of the natural logarithm:

```{r echo=2}
suppressMessages(library(cheby))
base <- d1.poly( fn=log, range=c(.01,4), iOrder=6, iPts=7 )
```

This computes a 6th-order approximation to `log` on the interval $[0.01,4]$ (recall that `log` is not continuous at zero) using 7 grid points (the minimum possible).   The output is a function that we can use straight away.  We can check the approximation at a few points:

```{r results='hold'}
c( base(1), log(1) )
c( base(2), log(2) )
```

So the approximation is good to around 1 decimal point near 0, and 2 decimal places for higher values of $x$.  Better still, we can see the approximation visually:

```{r fig.width=8, fig.height=4, echo=1:2, fig.cap = "Sixth-order approximation for log."}
plot( log, xlim=c(.01, 4), type='l', lwd=2, lty=2 ) 
plot( base, xlim=c(.01, 4), col='blue', add=TRUE, lwd=2 )
legend( 'bottomright', c('log', 'Order 6 polynomial approximation'), 
        col=c('black','blue'), lty=c(2,1), bty='n', lwd=2 )
```

So the approximation is very good for most of the approximation range, although given the discontinuity at zero it struggles for small $x$.  And it is fast:

```{r}
library(microbenchmark)
microbenchmark(base <- d1.poly( fn=log, range=c(.01,4), iOrder=6, iPts=7 ))
```

However, the approximation is not strictly concave - it oscilates slightly around the approximated function.  Note how the dashed log function shows up alternately above and below the approximation.  In some numerical calculations, preserving shape properties of the function can be important.  Later we will discuss how value function will often fail if the continuation value function is not concave, as then there is no unique local interior maximum.  This is not a problem which can be eliminated by simply increasing the order of the approximation, either.

## Extracting more information about the approximation

To extract more information about the approximation, set the `details` flag to `TRUE`.  

```{r}
base.deets <- d1.poly( fn=log, range=c(.01,4), iOrder=6, iPts=7, details=TRUE )
```

This returns a five-element list with elements:

* `fn` The polynomial approximation function.  So `base.deets$fn(1)$ will equal $base(1)$
* `poly` The polynomial representation of the Chebychev approximation.  This element is a polynomial from the `polynom` package, so can be manipulated easily (see the documentation of that package for more details).  It is also defined only on the $[-1,1]$ interval, so needs to be translated to the approximation interval to match the output of the `fn` element.
* `fn.deriv` The exact derivative of the polynomial approximation function.  Computed by differentiating the polynomial representation and translating to the approximation interval.  This is very useful for optimization problems where the derivatives are required.  Plotting this function (not shown here) will generate upward-sloping parts of the function, confirming that the approximation is not globally concave.  We can check also this calculation by comparing to the finite-difference calculation:
```{r results='hold'}
1e06 * ( base(1+1e-06) - base(1) )
base.deets$fn.deriv(1)
```
* `fn.deriv.poly`  The polynomial represenatation of the derivative.  The analogue of `poly`, but for `deriv`. 
* `residuals` The residuals of the approximation.  All zero if `iPts` is one more than `iOrder`.

## Pre-computing function values

The "standard" usage of `d1.poly` above integrates function calulcation and approximation.  This is convenient - it simply allows one to pass a function to the approximation and get a function back.  But it is not very flexible - it requires a delf-contained function to be evaluated seperately at each of the points in the grid.  In many applications this is not efficient.  For example, if the target function to be evaluated contains a computationally demanding common component, it may be quicker to calculate that on its own and then include that value in the evaluation of $f(\cdot)$ at each of the grid points.

Here is an example of how to use this syntax

```{r}
grid.pts <- d1.grid( c( .01, 4), 7 )
    # The grid of Chebychev nodes
log.vals <- sapply( grid.pts, log )
    # The target function evaluated at the approximation nodes
base.grid <- d1.poly( fn.vals=log.vals, grid=grid.pts, range=c(.01, 4), iOrder=6, iPts=7, details=TRUE )
```

The inclusion of the argument `grid` here is superfluous, as without this `d1.poly` assumes that the function values are evaluated on a grid of Chebychev nodes, which is exactly what `d1.grid` produces here.  If you wish to evaluate the function on a different grid of nodes, you should use this argument to specify that grid.

## Passing function parameters

One can pass function parameters using the argument `fn.opts`.  However, this *requires* that the target function have all its parameters enter through a single argument named `opts` (using a list to pass multiple parameters where required).  

For example, to compute a fifth-order approximation to the function $A k ^ \alpha$ where $A$ and $\alpha$ are parameters:
```{r fig.width=8, fig.height=4, fig.cap = "Approximating functions with parameters."}
target <- function( k, opts )
  return( opts$A  * k ^ opts$alpha ) 
apx.1 <- d1.poly( target, c(.01,2), 5, 6, fn.opts=list(A=1, alpha=.7) )
apx.2 <- d1.poly( target, c(.01,2), 5, 6, fn.opts=list(A=1.1, alpha=.4) )
plot( function(x) target(x, opts=list(A=1, alpha=.7) ), xlim=c(.01,2), lwd=2, lty=2, ylab='' )
plot( function(x) target(x, opts=list(A=1.1, alpha=.4) ), xlim=c(.01,2), lwd=2, lty=2, add=TRUE )
plot( apx.1, xlim=c(.01,2), col='blue', lwd=2, add=TRUE )
plot( apx.2, xlim=c(.01,2), col='blue', lwd=2, add=TRUE )
```


# One-dimensional shape-preserving Chebychev approximation

## The theory

[tbc - from Judd]

## Usage


# An example with applications in economics

## The problem

## Value function iteration

## Solving the first-order conditions directly

# Two-dimensional approximation

# References
Judd (1998), Numerical Methods in Economics, MIT Press


<!--Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` setion of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872)) -->